---
title: "Bayesian analysis of distance sampling data using the nimbleDistance package"
author: "Michael Scroggie"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_width: 5
    fig_height: 3.5
    number_sections: false
    toc: false
vignette: >
  %\VignetteIndexEntry{Bayesian analysis of distance sampling data using the nimbleDistance package}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Background

The *nimbleDistance* package provides functions for implementing distance-sampling analyses using *nimble* (de Valpine et al. 2017, 2020): a fast, extensible implementation of the BUGS language for analysing Bayesian statistical models. The package is at an early stage of development and currently includes functions for fitting half-normal and hazard-rate detection models to line-transect and point-count distance data. 

This vignette demonstrates some simple distance sampling analyses using *nimbleDistance*, and compares the results with those obtained using equivalent distance-detection models fit by maximum likelihood using package *Distance* (Miller et al. 2019). As an example, we analyse line-transect collected from a population of the Winter Wren *(Troglodytes troglodytes)* in Fife, Scotland. This dataset was originally described by Buckland et al. (2006). In addition, we fit point-distance models to point data for the Savannah Sparrow  *(Passerculus sandwichensis)* (Knopf et al. 1988). In both cases, we compare the estimated parameters of the distance-detection functions obtained using *nimbleDistance* to those estimated using *Distance*.

## Simple analysis of line transect distance sampling data

We begin by loading the required packages. The wren line-transect data are included in package *Distance*.

```{r packages_and_data}
library(Distance)
library(mrds)
library(nimbleDistance)
data(wren_lt)
```

First, we fit simple half-normal and hazard-rate distance models to the wren data using package *Distance*.
```{r simple_fit}
wren.hn <- ds(data=wren_lt, key="hn", adjustment=NULL)
wren.hr <- ds(data=wren_lt, key="hr", adjustment=NULL)
```

Plotting the fitted detection functions shows that the hazard-rate function fits somewhat better, reflecting the presence of a strong "shoulder" in the distribution of observed distances at approximately 75 meters from the transect lines. 
```{r plot_fits, fig.height=8, fig.width=5, fig.cap="Figure 1. Distance functions for the wren data fitted using package Distance"}
layout(matrix(1:2, nrow=2))
plot(wren.hn, main="Half-normal", las=1)
plot(wren.hr, main="Hazard rate", las=1)
```

Next, we reproduce this very simple analysis using Bayesian models implemented with the functions provided in the *nimbleDistance* package. Firstly, a half-normal function is fitted. The results show generally good agreement between the estimates of the scale parameter of the half normal function (sigma) and the fitted distance function agrees closely with that obtained using *Distance* above.

```{r nimble_halfnormal, fig.height=7, fig.width=7, fig.cap="Figure 2. Half-normal distance function fitted to the wren data using nimbleDistance. The blue vertical line on the histogram is the estimate of sigma from the analysis using Distance. For the distance function, the black line is the fitted function from Distance, the red line is the function fitted in nimbleDistance", cache=TRUE}
y<-wren_lt$distance
nind<-length(y)
distCodeV<-nimbleCode({
 y[1:nind]~dHN_V(sigma, 100, point=0)
 sigma~dunif(0, 100)
 })
#inits and monitors
inits <- function() list(sigma=50)
params <- c("sigma")

#generate some MCMC samples
samples <- nimbleMCMC(
code = distCodeV,
constants = list(y=y, nind=nind), ## the distance data
inits = inits,
monitors = params,
niter = 5000,
nburnin = 1000)

par(mar=c(4,4, 2, 1))
layout(matrix(1:4, nrow=2))
hist(samples[,"sigma"], col="red", breaks=100, las=1, xlab=expression(paste(sigma)), main="", border=NA)
abline(v=exp(wren.hn$ddf$par), col="blue", lwd=2) #Distance uses log-scale for parameters
#traceplots
plot(samples[,"sigma"], type="l", col="red", las=1, ylab="value", xlab="iteration")
#fitted distance function
plot(wren.hn, lwd=2, las=1)
lines(x=0:100, y=dHN(0:100, mean(samples[,"sigma"]), point=0)/
                 dHN(0, mean(samples[,"sigma"]), point=0),
        col="red")
```

Next, we fit the hazard-rate distance function to the wren data using *nimbleDistance* and compare the results with those obtained using *Distance*. Again, the estimates of the parameters (sigma and b) are in close agreement, and the fitted detection function closely resembles that obtained using *Distance*.

```{r nimble_hazard, fig.height=7, fig.width=7, fig.cap="Figure 3. Hazard-rate distance function fitted to the wren data using nimbleDistance. The blue vertical lines on the histograms are the estimates of sigma and b from the analysis using Distance. For the distance function, the black line is the fitted function from Distance, the red line is the function fitted in nimbleDistance", cache=TRUE}
y<-wren_lt$distance
nind<-length(y)
distCodeV<-nimbleCode({
 y[]~dHR_V(b, sigma, 100, point=0)
 sigma~dunif(1, 200)
 b~dunif(1, 50)
 })
#inits and monitors
inits <- function() list(sigma=50, b=5)
params <- c("sigma", "b")

#generate some MCMC samples
samples <- nimbleMCMC(
code = distCodeV,
constants = list(y=y, nind=nind), ## the distance data
inits = inits,
monitors = params,
niter = 5000,
nburnin = 1000)

par(mar=c(4,4, 2, 1))
layout(matrix(1:4, nrow=2))
#histograms
hist(samples[,"sigma"], col="red", breaks=100, xlab=expression(paste(sigma)), main="", border=NA, las=1)
abline(v=exp(wren.hr$ddf$par)[2], col="blue", lwd=2)
hist(samples[,"b"], col="red", xlab="b", breaks=100, main="", border=NA, las=1)
abline(v=exp(wren.hr$ddf$par)[1], col="blue", lwd=2)
#traceplots
plot(samples[,"sigma"], type="l", ylim=c(0, 100), col="red", las=1, ylab="value", xlab="iteration")
lines(samples[,"b"], col="red")
#fitted distance function
plot(wren.hr, lwd=2, las=1)
lines(x=0:100, y=dHR(0:100, mean(samples[,"b"]), mean(samples[,"sigma"]), point=0)/
        dHR(0, mean(samples[,"b"]), mean(samples[,"sigma"]), point=0),
      col="red")
```

## Simple analysis of point count distance sampling data

Next, we turn to analysis of data from point-transects. The Savannah Sparrow data set (Knopf et al. 1988) provided in the *Distance* package provides a useful example. First, we analyse this data set using package *Distance* to obtain a maximum likelihood estimate of the scale parameter (sigma) of the half-normal distance-detection function. 

```{r, fig.height=7, fig.width=7, fig.cap="Figure 4. Detection distances for Savannah Sparrow point surveys.", cache=TRUE}
data(Savannah_sparrow_1980)
hist(Savannah_sparrow_1980$distance, xlab="Distance (m)", 
		 main="Savannah sparrow point transects", breaks=12)

sasp.hn <- ds(data=Savannah_sparrow_1980, key="hn", adjustment=NULL,
							transect="point")

exp(sasp.hn$ddf$par)  #sigma =20.89
```

The analysis using *Distance* yields an estimate of $\hat\sigma =$ 20.89. 

Next, we analyse the same data using *nimbleDistance*. Note the use of the argument 'point' in the model code to indicate that the distance data were obtained using point surveys, rather than line transects. 
```{r}
y<-Savannah_sparrow_1980$distance
y<-y[!is.na(y)] #drop the NAs
nind<-length(y)
distCodeV<-nimbleCode({
	y[1:nind]~dHN_V(sigma, Xmax=80, point=1)
	sigma~dunif(0, 50)
})
#inits and monitors
inits <- function() list(sigma=5)
params <- c("sigma")
#generate some MCMC samples
samples <- nimbleMCMC(
	code = distCodeV,
	constants = list(y=y, nind=nind), ## the distance data
	inits = inits,
	monitors = params,
	niter = 5000,
	nburnin = 1000)

```

```{r, fig.height=7, fig.width=7, fig.cap="Figure 5. Parameter (sigma) for the half-normal distance function fitted to the savannah sparrow data using nimbleDistance. The histogram gives the distribution of MCMC samples. The blue vertical line is the posterior mean. The red vertical line is the maximum likelihood estimate of sigma obtained from an analysis of the same data using Distance.", cache=TRUE}
hist(samples[,"sigma"], breaks=30, xlab="sigma", main="")
abline(v=exp(sasp.hn$ddf$par), col="red", lwd=3) #MLE from Distance
abline(v=mean(samples[,"sigma"]), col="blue", lwd=3) #posterior mean from nimbleDistance
```

As expected, the posterior mean of $\sigma$ estimated using *nimbleDistance* is very close to the MLE estimated using *Distance*.

## References

Buckland, S. T. (2006) Point-transect surveys for songbirds: robust methodologies. *The Auk* 123: 345–357.

de Valpine, P., D. Turek, C.J. Paciorek, C. Anderson-Bergman, D. Temple Lang, and R. Bodik. 2017. Programming with models: writing statistical algorithms for general model structures with NIMBLE. *Journal of Computational and Graphical Statistics* 26: 403-413.

de Valpine P, Paciorek C, Turek D, Michaud N, Anderson-Bergman C, Obermeyer F, Wehrhahn Cortes C, Rodriguez A, Temple Lang D, Paganin S (2020). *NIMBLE: MCMC, Particle Filtering, and Programmable Hierarchical Modeling*. R package version 0.10.0

Knopf, F.L., J.A. Sedgwick, and R.W. Cannon. (1988) Guild structure of a riparian avifauna relative to seasonal cattle grazing. *Journal of Wildlife Management* 52: 280–290.

Miller DL, Rexstad E, Thomas L, Marshall L, Laake JL (2019). Distance Sampling in R. *Journal of Statistical Software*,
89: 1-28.

